Association between trauma severity models and opportunities for improvement: A retrospective cohort study

### 2022-09-12 studyprogressreport.Rmd
progress on method

Management of severe trauma is highly time-sensitive and is also dependent on the actions of a multidisciplinary chain of healthcare providers. Errors in management are a common cause of preventable deaths [@prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira]. Outcomes in severe trauma are dependent on the quality of care received, and it has been shown in the US that trauma quality improvement (QI) programs improve outcomes [@ustqi]. However, it has been shown that the implementation of QI programs differs widely between clinics [@tqipanglo] and there is still plenty of progress to be made in improving these programs.

@@

The gold standard of trauma quality improvement programs worldwide are multidisciplinary morbidity and mortality (M&M) conferences.  M&M conferences review selected patient cases and attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions. The incidence of OFI function in this manner as a proxy for incidence of suboptimal care. The WHO guidelines for trauma quality improvement programmes recommend regularly scheduled multidisciplinary conferences and give specific recommendations as to their structure and format [@whoguidelines]. Despite these  recommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqpianglo]. The WHO note themselves in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].

@@

In 2014 researchers in Norway released a new model, Norwegian Survival Prediction Model in Trauma (NORMIT), that attempts to address the weaknesses of TRISS and simultaneously provide a model based on Norwegian patient populations. NORMIT is based on the patient's age, New Injury Severity Score (NISS), comorbities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS. NISS differs from ISS in that NISS allows for scoring of multiple injuries within the same anatomical region. NORMIT also provides a method to apply a GCS score for intubated patients when calculating RTS.

@@

<!-- Present key elements of study design -->
This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH): the trauma registry and the trauma care quality database (TCQD). The TCQD is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews. These two databases will be linked and then analysis will be performed to estimate the association between the common TSS models TRISS and NORMIT <!-- And something else? --> and OFI using logistic regression. The predictive performance of these models will be evaluated using measures of discrimination and calibration.

@@

KUH is located in Stockholm, Sweden. KUH's trauma centre is equivalent to a level 1 trauma centre according to American College of Surgeons standards [@nksstats]. KUH is the primary trauma centre for the region of Stockholm. In addition, KUH has agreements with several other regions to provide trauma care for patients with severe trauma. KUH is thereby the major trauma centre for a population of almost 3 million people [@nksstats].

@@

The trauma care quality register is an internal hospital register and registers the outcome of M&M conferences. Each case reviewed is registered in the trauma care quality register with data about the outcome of the review.

@@

Participants include all patients registered in both the trauma registry and the TCQD at KUH. <!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) -->

@@

### Exposures
The primary exposures of interest are trauma severity scores according to the TRISS and NORMIT models. Both of these models estimate the Ps based on expected mortality within 30 days. This is expressed a percentage and is a continuous variable. The variables integrated into these models are taken from arrival at hospital.

@@

The NORMIT model incorporates the patient's age, New Injury Severity Score (NISS), comorbities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS. The NISS differs from the ISS in that the NISS is the sum of the squares from only the three most injured anatomical regions.

@@

Since all eligible participants must exist in both the trauma and the trauma care quality register, and the trauma care quality registry is a subset of the trauma registry, the number of participants is limited to the number of cases registered in the trauma care quality database. The trauma care quality database includes around 2200 patients who were selected for review between 2014 and 2021.

<!-- Update these numbers later when I have real numbers -->

Quantitative variables
----------------------
<!--Om du behöver "skapa" ny data genom att manipulera den insamlade (tvek i ditt fall) så ska du beskriva det här. Annars kan detta stycke slås ihop med variables ovan. -->
<!-- Explain how quantitative variables were handled in the
analyses. If applicable, describe which groupings were chosen and why
-->

### Exposures

<!--Om du inte tydligt definerat modellernas variabler behöver du göra det här, alltså för exv triss som är GCS, RR, BT + ISS osv osv-->
<!-- Really not sure exactly what to put here
-->

@@

R, a programming language for statistical computing, will be used for compilation and analysis of data. Conversion and handling of variables will be carried out according to the SweTrau manual. Results will be presented with a confidence interval of 95%. A p-value of <0.05 will be considered significant. 
<!-- Jo det kommer nog bli ccs, du kommer exkludera alla patienter som delsa saknar outcome, OFI alternativt där det saknas nog data för att bygga olika scores. -->
<!-- Complete case study? -->
<!--

Should answer these questions, where applicable:

* All statistical methods, incl. those used to control for confounding
* Any methods to examine subgroups and interactions
* How was missing data addressed?
* Cohort study: If applicable, explain how loss to follow-up was addressed 
* Any sensitivity analyses? -->

Ethical considerations
----------------------

>>>>

Management of severe trauma is highly time-sensitive and is also dependent on the actions of a multidisciplinary chain of healthcare providers. Errors in management are a common cause of preventable deaths [@prevdeathsghorbani; @prevdeathsmontmany; @prevdeathsteixeira]. Outcomes in severe trauma are dependent on the quality of care received, and it has been shown in the US that trauma quality improvement (QI) programs improve outcomes [@ustqi]. However, it has also been shown that the implementation of QI programs differs widely between clinics [@tqipanglo] and there is still progress to be made.

@@

The gold standard of trauma quality improvement programs worldwide are multidisciplinary morbidity and mortality (M&M) conferences.  M&M conferences review selected patient cases and attempt to establish whether there were opportunities for improvement (OFI) in each specific case, and, where necessary, implement corrective actions. The incidence of OFI function in this manner as a proxy for incidence of suboptimal care. The WHO guidelines for trauma quality improvement programmes recommend regularly scheduled multidisciplinary conferences and give specific recommendations as to their structure and format [@whoguidelines]. Despite these  recommendations, there is a still a great degree of variation amongst clinics in how M&M conferences are carried out, with a significant number of clinics still using an informal and unstructured format [@tqipanglo]. The WHO note themselves in their guidelines that M&M conferences are not used to their fullest potential globally [@whoguidelines].

@@

In 2014 researchers in Norway released a new model, Norwegian Survival Prediction Model in Trauma (NORMIT), that attempts to address the weaknesses of TRISS and simultaneously provide a model based on Norwegian patient populations. NORMIT is based on the patient's age, New Injury Severity Score (NISS), comorbidities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS. NISS differs from ISS in that NISS allows for scoring of multiple injuries within the same anatomical region. NORMIT also provides a method to apply a GCS score for intubated patients when calculating RTS.

@@

<!-- Present key elements of study design -->
This is a single-centre retrospective cohort study, using data from two registries at Karolinska University Hospital (KUH): the trauma registry and the trauma care quality database (TCQD). The TCQD is a subset of the trauma registry and contains data on trauma patients who were selected for M&M reviews. These two databases will be linked and then analysed to estimate the association between the common TSS models TRISS and NORMIT and OFI using logistic regression. The predictive performance of these models will be evaluated using measures of discrimination and calibration.

@@

KUH is located in Stockholm, Sweden. KUH's trauma centre is equivalent to a level 1 trauma centre according to American College of Surgeons standards [@nksstats]. KUH is the primary trauma centre for the region of Stockholm and also has agreements with several other regions to provide trauma care for patients with severe trauma. KUH is thereby the major trauma centre for a population of almost 3 million people [@nksstats].

@@

The TCQD is an internal hospital register and registers the outcome of M&M conferences. Each case reviewed is registered in the trauma care quality register with data about the outcome of the review.

@@

Participants include all patients registered in both the trauma registry and the TCQD at KUH. <!--När du kommit igång med analysen så ska faktiskt nummer in här (yes ska göra det) --> Furthermore, patients <15 years of age will be excluded as paediatric trauma is a seperate clinical entity.

@@

### Exposures
The primary exposures of interest are trauma severity scores according to the TRISS and NORMIT models. Both of these models estimate the Ps. This is expressed a probability and is a continuous variable. The variables integrated into these models are taken from arrival at hospital.

@@

The NORMIT model incorporates the patient's age, New Injury Severity Score (NISS), comorbidities prior to traumatic injury based on the ASA physical status classification system (ASA-PS), and their RTS.

@@

Since all eligible participants must exist in both the trauma and the trauma care quality register, and the trauma care quality registry is a subset of the trauma registry, the number of participants is limited to the number of cases registered in the trauma care quality database. The trauma care quality database includes around 2200 patients who were selected for review between 2014 and 2021. The exact study size has not yet been established at the time of writing this report.

Quantitative variables
----------------------

The NORMIT and TRISS scores will be calculated using the following variables from the trauma registry database: ISS, NISS, RTS, GCS, systolic blood pressure, respiratory rate, ASA, and age.

@@

R, a programming language for statistical computing, will be used for compilation and analysis of data. Conversion and handling of variables will be carried out according to the SweTrau manual. This is a complete case study. In cases where data necessary for analysis is missing, i.e. the variables referred to above, the participant will be excluded from the study. Results will be presented with a confidence interval of 95%. A p-value of <0.05 will be considered significant.

Ethical considerations
======================
